Meta analysis of differential expression results from AMP-AD (Control vs AD, reprocessed RNASeq)

### 2017-03-22 code/Rmd/MetaAnalysis.Rmd
temp commit

### Get covariates/sample metadata from synapse
```{r covariates}
covariates.ids = c(ROSMAP = 'syn8018344',
                   MSBB = 'syn8073656',
                   MAYO = 'syn8028561')
covariates = lapply(covariates.ids, function(id){
  x = downloadFile(id)
  colnames(x)[1] = 'SampleID'
  return(x)
})

@@

covariates$MSBB = dplyr::filter(covariates$MSBB, BrodmannArea.Dx %in% c('BM10.SD', 'BM22.SD', 'BM36.SD', 
                                                                        'BM44.SD', 'BM36.ND', 'BM10.ND', 
                                                                        'BM44.ND', 'BM22.ND')) %>%
  dplyr::select(SampleID, BrodmannArea.Dx) %>%
  tidyr::separate(BrodmannArea.Dx, c('BrainRegion', 'Dx')) %>%
  dplyr::mutate(Study = 'MSSM',
                Dx = factor(Dx, levels = c('ND','SD')),
                BrainRegion = factor(BrainRegion, labels = c('FP', 'STG', 'PHG', 'IFG'))) %>%
  dplyr::select(SampleID, Study, BrainRegion, Dx)

@@

### Get differential expression data from synapse
```{r diffexp}
diff.exp.ids = c(ROSMAP = 'syn8018358',
                 MSBB = 'syn8073678',
                 MAYO = 'syn8028572')
diff.exp = lapply(diff.exp.ids, downloadFile)

@@

pl = list()
for(region in c('ROSMAP', 'MSBB', 'MAYO')){
  p = ggplot(tmp %>% filter(Study == region), aes(x = logFC, y = -log10(adj.P.Val), color = Direction)) + geom_point()
  p = p + scale_color_manual(values = c('green','grey','red'))
  p = p + facet_grid(Study~.+Region) + theme(legend.position = 'top')
  pl[[region]] = p
}
multiplot(plotlist = pl, cols = 1)
```

@@

### Get hgnc_symbol mapping from biomaRt
```{r biomart}
# Define biomart object
mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", dataset = "hsapiens_gene_ensembl", host = "dec2016.archive.ensembl.org")

# Get background gene ids
backgroundGenes = sapply(diff.exp, function(x) x$ensembl_gene_id) %>%
  Reduce(c,.)

# Query biomart
Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol"),
                       filters = "ensembl_gene_id", values = backgroundGenes,
                       mart = mart)

Entrez2Ensembl <- as.data.frame(org.Hs.egENSEMBL2EG)
Entrez2HGNC <- as.data.frame(org.Hs.egSYMBOL2EG)
HGNC2Ensembl <- left_join(Entrez2HGNC, Entrez2Ensembl)
HGNC2Ensembl = HGNC2Ensembl[grep('^LOC',HGNC2Ensembl$symbol),c('ensembl_id', 'symbol')] %>%
  filter(!is.na(ensembl_id), !is.na(symbol)) %>%
  dplyr::rename(ensembl_gene_id = ensembl_id, hgnc_symbol = symbol)

Ensemble2HGNC = rbindlist( list(Ensemble2HGNC, HGNC2Ensembl), use.names = T, fill = T) %>%
  unique
```

### Consistently differentially expressed genes (in all 7 brain regions)
```{r diffexp.all, fig.height=16, fig.width=10}
# Find genes that are differentially expressed in each brain region
diff.exp.gl = lapply(diff.exp, function(x){
  dplyr::filter(x, abs(logFC) >= log2(1.2) & adj.P.Val <= 0.05 & sign(CI.L) == sign(CI.R)) %>%
    dplyr::select(Region, gene.id) %>%
    unique
}) %>%
  rbindlist(use.names = T, fill = T) %>%
  dlply(.(Region), .fun = function(x){x$gene.id})

tmp1 = rbindlist(diff.exp, use.names = T, fill = T) %>%
  dplyr::filter(gene.id %in% Reduce(c, diff.exp.gl)) %>%
  dplyr::select(-hgnc_symbol) %>%
  left_join(Ensemble2HGNC)
tmp1$hgnc_symbol[tmp1$ensembl_gene_id == 'ENSG00000275830'] = 'LOC101927646'

@@

tmp = tmp1 %>%
  dplyr::filter(hgnc_symbol != '') %>%
  group_by(hgnc_symbol) %>%
  summarise(count = sum(Direction != 'NONE')) %>%
  filter(count >= 5) 
writeLines(paste('Differentially expressed genes in any 5 brain regions at an FDR of 0.05 and fold change of 1.2 are',
                 length(unique(tmp$hgnc_symbol))))
kable(data.frame(ngenes = length(unique(tmp$hgnc_symbol)),
                 hgnc_symbol = tmp$hgnc_symbol %>% unique %>% paste(collapse = ',')))

@@

pl = list()
p = ggplot(tmp1 %>% 
             filter(hgnc_symbol %in% tmp.eff.sz$hgnc_symbol[tmp.eff.sz$es>2.6]), 
           aes(x = Region, y = logFC, color = Study, fill = Direction)) + geom_bar(stat = 'identity')
p = p + facet_grid(hgnc_symbol~.) + geom_errorbar(aes(ymin=CI.L, ymax=CI.R), width = 0.2)
p = p + theme(legend.position = 'top') + scale_fill_manual(values = c('grey', 'red'))
pl[[1]] = p

p = ggplot(tmp1 %>% 
             filter(hgnc_symbol %in% tmp.eff.sz$hgnc_symbol[tmp.eff.sz$es<2.6]), 
           aes(x = Region, y = logFC, color = Study, fill = Direction)) + geom_bar(stat = 'identity')
p = p + facet_grid(hgnc_symbol~.) + geom_errorbar(aes(ymin=CI.L, ymax=CI.R), width = 0.2)
p = p + theme(legend.position = 'top')  + scale_fill_manual(values = c('green','grey','red'))
pl[[2]] = p

@@

### Pathway enrichment analysis
```{r pathway.enrich}
setsToTest = rbindlist(diff.exp, use.names = T, fill = T) %>%
  filter(adj.P.Val <= 0.05, Direction != 'NONE') %>%
  dlply(.(Region, Direction), .fun = function(x){
    unique(x$hgnc_symbol)
  })

@@

# Write results to file
tmp1 = rbindlist(enrichResults, use.names = TRUE, idcol = 'ComparisonName', fill = TRUE) %>%
  filter(!is.na(Category))

>>>>

### Get covariates/sample metadata from synapse
```{r covariates}
covariates.ids = c(ROSMAP = 'syn8456631',
                   MSBB = 'syn8484996',
                   MAYO = 'syn8466814')
covariates = lapply(covariates.ids, downloadFile)

@@

covariates$MSBB = covariates$MSBB %>%
  tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion', 'Diagnosis'), sep = '\\.') %>%
  dplyr::filter(Diagnosis %in% c('SD', 'ND')) %>%
  dplyr::mutate(Study = 'MSSM',
                Dx = factor(Diagnosis, levels = c('ND','SD')),
                BrainRegion = factor(BrainRegion, labels = c('FP', 'STG', 'PHG', 'IFG'))) %>%
  dplyr::select(SampleID, Study, BrainRegion, Dx)

@@

### Get differential expression data from synapse
```{r diffexp}
diff.exp.ids = c(ROSMAP = 'syn8456721',
                 MSBB = 'syn8502894',
                 MAYO = 'syn8468023')
diff.exp = lapply(diff.exp.ids, downloadFile)

@@

pl = list()
for(region in c('ROSMAP', 'MSSM', 'MAYO')){
  p = ggplot(tmp %>% filter(Study == region), aes(x = logFC, y = -log10(adj.P.Val), color = Direction)) + geom_point()
  p = p + scale_color_manual(values = c('green','grey','red'))
  p = p + facet_grid(Study~.+Region) + theme(legend.position = 'top')
  pl[[region]] = p
}
multiplot(plotlist = pl, cols = 1)
```

@@

### Get hgnc_symbol mapping from biomaRt
```{r biomart}
# Get background gene ids
backgroundGenes = sapply(diff.exp, function(x) x$hgnc_symbol) %>%
  Reduce(c,.) %>% unique() %>% setdiff(c(''))
```

### Consistently differentially expressed genes (in all 7 brain regions)
```{r diffexp.all, fig.height=16, fig.width=10}
# Find genes that are differentially expressed in each brain region
diff.exp.gl = lapply(diff.exp, function(x){
  dplyr::filter(x, abs(logFC) >= log2(1.2) & adj.P.Val <= 0.05 & sign(CI.L) == sign(CI.R)) %>%
    dplyr::select(Region, ensembl_gene_id) %>%
    unique
}) %>%
  rbindlist(use.names = T, fill = T) %>%
  dlply(.(Region), .fun = function(x){setdiff(unique(x$ensembl_gene_id),c(''))})

tmp1 = rbindlist(diff.exp, use.names = T, fill = T) %>%
  dplyr::filter(ensembl_gene_id %in% Reduce(c, diff.exp.gl))

@@

tmp = tmp1 %>%
  group_by(ensembl_gene_id) %>%
  summarise(count = sum(Direction != 'NONE')) %>%
  filter(count == 7) %>%
  left_join(dplyr::select(tmp1, ensembl_gene_id, hgnc_symbol) %>% unique)

writeLines(paste('Differentially expressed genes in all 7 brain regions at an FDR of 0.05 and fold change of 1.2 are',
                 length(unique(tmp$hgnc_symbol))))
kable(data.frame(ngenes = length(unique(tmp$hgnc_symbol)),
                 hgnc_symbol = tmp$hgnc_symbol %>% unique %>% paste(collapse = ',')))

@@

pl = list()
p = ggplot(tmp1 %>% 
             filter(hgnc_symbol %in% c('MEIS3', 'PCSK1', 'ADAMTS2')), 
           aes(x = Region, y = logFC, color = Study, fill = Direction)) + geom_bar(stat = 'identity')
p = p + facet_grid(hgnc_symbol~.) + geom_errorbar(aes(ymin=CI.L, ymax=CI.R), width = 0.2)
p = p + theme(legend.position = 'top') + scale_fill_manual(values = c('green', 'red'))
pl[[1]] = p

p = ggplot(tmp1 %>% 
             filter(hgnc_symbol %in% c('FOXJ1', 'PRELP', 'S100A4')), 
           aes(x = Region, y = logFC, color = Study, fill = Direction)) + geom_bar(stat = 'identity')
p = p + facet_grid(hgnc_symbol~.) + geom_errorbar(aes(ymin=CI.L, ymax=CI.R), width = 0.2)
p = p + theme(legend.position = 'top')  + scale_fill_manual(values = c('green','grey', 'red'))
pl[[2]] = p

@@

### Pathway enrichment analysis
```{r pathway.enrich}
setsToTest = rbindlist(diff.exp, use.names = T, fill = T) %>%
  filter(adj.P.Val <= 0.05, Direction != 'NONE') %>%
  dlply(.(Region, Direction), .fun = function(x){
    setdiff(unique(x$hgnc_symbol),c(''))
  })

@@

# Write results to file
tmp1 = rbindlist(enrichResults, use.names = TRUE, idcol = 'ComparisonName', fill = TRUE) %>%
  filter(!is.na(Category)) %>%
  filter(fdr <= 0.01)

tmp2 = dplyr::select(tmp1, ComparisonName, GeneSetName, Odds.Ratio) %>%
  tidyr::spread(ComparisonName,Odds.Ratio)
rownames(tmp2) = tmp2[,1]
